Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-08-29 pkg/disttree/inst/draft/draft.Rnw
highlighted changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2268 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 394c73009df503993cb104295fbe472d06896c85

%% tikz
\usepackage{array,makecell,tikz}
\usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}

@@

# predictions
pred1 <- c(Axams_24to4$testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
pred2 <- c(Axams_24to4$testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
pred3 <- c(Axams_24to4$testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
pred4 <- c(Axams_24to4$testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))

@@

# plot predictions
points(x = pred1[1], y = pred1[2], col = pal[1], pch = 4)
points(x = pred2[1], y = pred2[2], col = pal[3], pch = 4)
points(x = pred3[1], y = pred3[2], col = pal[2], pch = 4)
points(x = pred4[1], y = pred4[2], col = pal[4], pch = 4)

lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}Total precipitation predictions by a distributional forest
at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
\end{figure}

@@

For obtaining probabilistic predictions from the tree for a (possibly new) set
of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
be ``sent down'' the tree and the corresponding segment-specific MLE has to be
obtained. This can also be understood as a weighted MLE where the weights
select those observations from the learning sample that fall into the same
segment:
$$
w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
$$
where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
for a given $\bold{z}$ is then fully specified by the estimated parameter
$\bm{\hat \theta}(\bold{z})$ where
$$
\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
$$

@@

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Bootstrap
  sampling is employed for each tree using a third of the predictors
  in each split of the tree (``mtry''). Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}.
  % stopping criteria:
  % tree_minsplit = 50, tree_minbucket = 20, tree_mincrit = 0.95,
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

@@

To assess the predictive performance, a full cross-validation is carried out
rather than relying on just the one fixed test set for the years 2009--2012.
To do so, a 10 times 7-fold cross-validation is carried out where each
splits the available 28~years into 7~subsets of 4~randomly selected years.
The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
(= 4~years) using the average CRPS across all observations.
The resulting 10~CRPS skill scores are displayed by boxplots in
Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
of around 4~percent, the distributional forest has a slightly higher improvement
of around 5.5~percent in median.

@@

These hypotheses are assessed with permutation tests based on
the multivariate linear statistic
$$
T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\bm{\hat{\theta}}, Y_i)\right)
$$
where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
type of the transformation function $v_l$ depends on the type of the split 
variable $Z_l$. If $Z_l$ is numeric then $v_l$ is simply the identity 
function $v_l(Z_{li}) = Z_{li}$ and therefore $T_l \in \mathbb{R}^k$ as 
the ``vec'' operator converts the $1 \times k$ matrix into a $k$ column vector. 
If $Z_l$ is a categorical variable with $H$ categories then 
$v_l(Z_{li}) = e_H(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$ 
such that $v_l$ is a $H$-dimensional unit vector where the element corresponding to 
the value of $Z_{li}$ is $1$. In this case the statistic $T_l \in \mathbb{R}^{H \cdot k}$
as the ``vec'' operator converts the $H \times k$ matrix into a $k$ column vector 
by column-wise combination. Observations with missing values are excluded from the sums.

With the conditional expectation $\mu_l$ and the covariance 
$\Sigma_l$ of $T_l$ as derived by 
\cite{Strasser+Weber:1999} the test statistic can be standardized. 
The observed multivariate linear statistic $t_l$ which is either an
$k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of Y,
is mapped onto the real line by a univariate test statistic $c$. 
In the application of this paper a quadratic form is chosen, such that
$$
c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
$$
where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$. Alternatively,
the maximum of the absolute values of the standardized linear statistic
can be considered.

The smaller the $p$-value corresponding to the standardized test 
statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
from the assumption of independence between the scores and the split 
variable $Z_l$.
%
After Bonferroni-adjusting the $p$-values it has to be assessed whether
any of the resulting $p$-values is beneath the selected significance level. 
If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
is chosen as splitting variable. Otherwise no further split is made is in this node.

The breakpoint that leads to the highest discrepancy between score functions 
in the two resulting subgroups is selected as split point. 
This is measured by the linear statistic
$$
T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\bm{\hat{\theta}}, Y_i)
$$
for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
two new subgroups, without any particular ordering, that are defined by 
splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
The split point is then chosen as follows:
$$
r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
$$
One repeats the testing and splitting procedure in each of the resulting 
subgroups until some stopping criterion is reached. This criterion 
can for example be a minimal number of observations in a node or a 
minimal $p$-value for the statistical tests. In that way pre-pruning
is applied in order to find right-sized trees and hence 
avoid overfitting.

>>>>

%% tikz
\usepackage{array,makecell,tikz,color,soul}
\usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}

@@

# observations
obs1 <- c(Axams_24to4$testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
obs2 <- c(Axams_24to4$testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
obs3 <- c(Axams_24to4$testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
obs4 <- c(Axams_24to4$testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))

@@

# plot observations
points(x = obs1[1], y = obs1[2], col = pal[1], pch = 4, cex = 1.4)
points(x = obs2[1], y = obs2[2], col = pal[3], pch = 4, cex = 1.4)
points(x = obs3[1], y = obs3[2], col = pal[2], pch = 4, cex = 1.4)
points(x = obs4[1], y = obs4[2], col = pal[4], pch = 4, cex = 1.4)

lines(x = c(obs1[1], obs1[1]), y = c(obs1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(obs2[1], obs2[1]), y = c(obs2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(obs3[1], obs3[1]), y = c(obs3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(obs4[1], obs4[1]), y = c(obs4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}\hl{Total precipitation predictions by a distributional forest
at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}}
\end{figure}

@@

For obtaining probabilistic predictions from the tree for a (possibly new) set
of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
be ``sent down'' the tree and the corresponding segment-specific MLE has to be
obtained. This can also be understood as a weighted MLE where the weights
select those observations from the learning sample that fall into the same
segment:
$$
w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
$$
where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
for a given $\bold{z}$ is then fully specified by the estimated parameter
$\bm{\hat \theta}(\bold{z})$ where
$$
\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
$$
\hl{Even though this parameter depends on the given $z$ it has to be noted that
it is the same for all other observations that are in the same terminal node of 
the tree as $z$ and therefore does not necessarily have to be recalculated for
each of them.}

@@

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Bootstrap
  sampling is employed for each tree using a third of the predictors
  in each split of the tree (\code{mtry}). Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}. 
  \hl{As stopping criteria the minimal number
  of observations to perform a split} (\code{minsplit}) \hl{is set to 50, 
  the minimal number of observations in a terminal node} (\code{minbucket})
  \hl{is set to 20 and the significance level for variable selection} (\code{alpha}) 
  \hl{is kept at the default value $1$.}
  % stopping criteria:
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

  % default in ctree_control within distforest:
  % mincriterion = 0

  % default values in ctree_control:
  % alpha = 0.05, mincriterion = 1 - alpha,
  % minsplit = 20L, minbucket = 7L,
  % minprob = 0.01

@@

To assess the predictive performance, a full cross-validation is carried out
rather than relying on just the one fixed test set for the years 2009--2012.
\hl{To do so, a 10 times 7-fold cross-validation is carried out where each
splits the available 28~years into 7~subsets of 4~randomly selected and 
therefore mostly not subsequent years.}
The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
(= 4~years) using the average CRPS across all observations.
The resulting 10~CRPS skill scores are displayed by boxplots in
Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
of around 4~percent, the distributional forest has a slightly higher improvement
of around 5.5~percent in median.

@@

These hypotheses are assessed with permutation tests based on
the multivariate linear statistic
$$
T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\bm{\hat{\theta}}, Y_i)\right)
$$
\hl{where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
type of the transformation function $v_l$ depends on the type of the split 
variable $Z_l$. If $Z_l$ is numeric then $v_l$ is simply the identity 
function $v_l(Z_{li}) = Z_{li}$ and therefore $T_l \in \mathbb{R}^k$ as 
the ``vec'' operator converts the $1 \times k$ matrix into a $k$ column vector. 
If $Z_l$ is a categorical variable with $H$ categories then 
$v_l(Z_{li}) = e_H(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$ 
such that $v_l$ is a $H$-dimensional unit vector where the element corresponding to 
the value of $Z_{li}$ is $1$. In this case the statistic $T_l \in \mathbb{R}^{H \cdot k}$
as the ``vec'' operator converts the $H \times k$ matrix into a $k$ column vector 
by column-wise combination.} Observations with missing values are excluded from the sums.

With the conditional expectation $\mu_l$ and the covariance 
$\Sigma_l$ of $T_l$ as derived by 
\cite{Strasser+Weber:1999} the test statistic can be standardized. 
\hl{The observed multivariate linear statistic $t_l$ which is either an
$k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of Y,
is mapped onto the real line by a univariate test statistic $c$.
In the application of this paper a quadratic form is chosen, such that}
$$
c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
$$
\hl{where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.} Alternatively,
the maximum of the absolute values of the standardized linear statistic
can be considered.

The smaller the $p$-value corresponding to the standardized test 
statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
from the assumption of independence between the scores and the split 
variable $Z_l$.
%
After Bonferroni-adjusting the $p$-values it has to be assessed whether
any of the resulting $p$-values is beneath the selected significance level. 
If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
is chosen as splitting variable. \hl{Otherwise no further split is made is in this node.}

The breakpoint that leads to the highest discrepancy between score functions 
in the two resulting subgroups is selected as split point. 
This is measured by the linear statistic
$$
T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\bm{\hat{\theta}}, Y_i)
$$
\hl{for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
two new subgroups, without any particular ordering, that are defined by 
splitting in split point $r$ of variable $Z_{l^{\ast}}$.
The split point is then chosen as follows:}
$$
r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
$$
One repeats the testing and splitting procedure in each of the resulting 
subgroups until some stopping criterion is reached. This criterion 
can for example be a minimal number of observations in a node or a 
minimal $p$-value for the statistical tests. In that way pre-pruning
is applied in order to find right-sized trees and hence 
avoid overfitting.
